A quantum parallel Markov chain Monte Carlo

### 2021-11-08 main.tex
quantum intro done

Quantum logic gates perform linear operations on qubits like $\ppsi$ and take the form of unitary matrices $\U$ satisfying $\U^\dagger \U=\I$ for $\U^\dagger$ the conjugate transpose of $\U$. One terrifically important single-qubit quantum gate is the \emph{Hadamard gate}
\begin{align}
\H = \frac{1}{\sqrt{2}} \left[\begin{array}{rr}
1 & 1 \\
1 & -1 \end{array}\right] \, .
\end{align}
The Reader may verify that $\H$ is indeed unitary and that its action maps $\ket{0}$ to $\ket{+}$ and $\ket{1}$ to $\ket{-}$.  Uncountably many single-qubit quantum gates exist, and some of them are useful. But the real power of quantum computing stems from the development of complex quantum gates that act on multiple qubits simultaneously.   To access this power, we need one more tool that appears in the statistics literature, albeit less frequently.  The Kronecker product between an $L$-by-$M$ matrix $\A$ and an $N$-by-$O$ matrix $\B$ is the $LN$-by-$MO$ matrix
\begin{align}\label{eq:kron}
\A \otimes \B = \begin{bmatrix}
A_{11} \B & \dots & A_{1M} \B \\
\vdots & \ddots & \vdots \\
A_{L1}\B & \cdots & A_{LM}  \B
\end{bmatrix} \, .
\end{align} 
In statistics, the Kronecker product features in the definition of a matrix normal distribution and is sometimes helpful when specifying the kernel function of a multivariate Gaussian process \citep{werner2008estimation}. Here, the product is equivalent to the parallel action of individual quantum gates on individual qubits.  Simple application of Formula \eqref{eq:kron} shows that
\begin{align}\label{eq:twoqubits}
\H \otimes \H = \frac{1}{2} \left[\begin{array}{rrrr}
1 & 1 & 1 & 1 \\
1 &  -1& 1 & -1 \\
1 & 1 & -1 & -1 \\
1 & -1 & -1 & 1
\end{array}\right] \quad \mbox{and} \quad \ket{0}\otimes\ket{0}=\begin{bmatrix}
1 \\ 0
\end{bmatrix} \otimes \begin{bmatrix}
1 \\ 0
\end{bmatrix} = \begin{bmatrix}
1 \\ 0 \\ 0 \\ 0
\end{bmatrix} \, .
\end{align} 
We may also denote the left product $\H^{\otimes2}$ and the right product $\ket{0}\ket{0}$ or $\ket{00}$. Letting $\ket{10}$, $\ket{01}$ and $\ket{11}$ take on analogous meanings, an immediate result of Equation \eqref{eq:twoqubits} is that
\begin{align}\label{eq:twodmult}
\H^{\otimes2} \ket{00} = \frac{1}{2}\Big(\ket{00} + \ket{01} + \ket{10} + \ket{11}\Big) \, .
\end{align}
The action of $\H^{\otimes2}$ transforms $\ket{00}$ into a superposition of the states $\ket{00}$, $\ket{10}$, $\ket{01}$ and $\ket{11}$, where the probability of selecting each is a uniform $(1/2)^2=1/4$. Writing so many $0$s and $1$s can be tedious, so an alternative notation becomes preferable. Exchanging $\ket{0}$ for $\ket{00}$, $\ket{1}$ for $\ket{01}$, $\ket{2}$ for $\ket{10}$, and $\ket{3}$ for $\ket{11}$, Equation \eqref{eq:twodmult} becomes the slightly more succinct
\begin{align}
\H^{\otimes2} \ket{0} = \frac{1}{2}\Big(\ket{0} + \ket{1} + \ket{2} + \ket{3}\Big) \, .
\end{align}
This formula extends generally to operations over $n$ qubits.  Now letting $N=2^n$, we have
\begin{align}
\H^{\otimes n} \ket{0} = \frac{1}{\sqrt{N}}\Big(\ket{0} + \ket{1} + \dots + \ket{N-1}\Big) =  \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \, .
\end{align}

@@

\section{Quantum parallelism, search and minimization}

>>>>

Quantum logic gates perform linear operations on qubits like $\ppsi$ and take the form of unitary matrices $\U$ satisfying $\U^\dagger \U=\I$ for $\U^\dagger$ the conjugate transpose of $\U$. One terrifically important single-qubit quantum gate is the \emph{Hadamard gate}
\begin{align}
\H = \frac{1}{\sqrt{2}} \left[\begin{array}{rr}
1 & 1 \\
1 & -1 \end{array}\right] \, .
\end{align}
The Reader may verify that $\H$ is indeed unitary and that its action maps $\ket{0}$ to $\ket{+}$ and $\ket{1}$ to $\ket{-}$.  In fact, the reverse is also true on account of the symmetry of $\H$.  The Hadamard gate thus takes the computational basis states in and out of superposition, facilitating a phenomenon called \emph{quantum parallelism}. Given a function $f: \{0,1\} \rightarrow \{0,1\}$, consider the two-qubit quantum \emph{oracle} gate $\U_f$ which takes $\ket{x}\ket{y}$ as input and returns $\ket{x}\ket{y\oplus f(x)}$ as output, where $\oplus$ denotes addition modulo 2.  Importantly, the output simplifies to $\ket{x}\ket{f(x)}$ for $y=0$. We now consider a quantum circuit that acts on two qubits by first applying the Hadamard transform $\H$ to the first qubit and then applying the oracle gate $\U_f$ to both. Using the state $\ket{0}\ket{0}$ as input, we have
\begin{align}\label{eq:quantparallel}
\ket{0}\ket{0} \longrightarrow \frac{1}{\sqrt{2}} \ket{0}\ket{0} + \frac{1}{\sqrt{2}} \ket{1}\ket{0}  \longrightarrow \frac{1}{\sqrt{2}} \ket{0}\ket{f(0)} + \frac{1}{\sqrt{2}} \ket{1}\ket{f(1)} \, .
\end{align}
The quantum circuit evaluates $f(\cdot)$ simultaneously over both inputs!  Unfortunately, the scientist who implements this circuit cannot directly access $\U_f$'s output, and measurement will provide only $\ket{0}\ket{f(0)}$ or $\ket{1}\ket{f(1)}$ with probability $1/2$ each.  Unlocking the potential of quantum parallelism will require a bit of ingenuity indeed.

Uncountably many single- and two-qubit quantum gates exist, but the real power of quantum computing stems from the development of complex quantum gates that act on multiple qubits simultaneously.   To access this power, we need one more tool that appears in the statistics literature, albeit less frequently.  The Kronecker product between an $L$-by-$M$ matrix $\A$ and an $N$-by-$O$ matrix $\B$ is the $LN$-by-$MO$ matrix
\begin{align}\label{eq:kron}
\A \otimes \B = \begin{bmatrix}
A_{11} \B & \dots & A_{1M} \B \\
\vdots & \ddots & \vdots \\
A_{L1}\B & \cdots & A_{LM}  \B
\end{bmatrix} \, .
\end{align} 
In statistics, the Kronecker product features in the definition of a matrix normal distribution and is sometimes helpful when specifying the kernel function of a multivariate Gaussian process \citep{werner2008estimation}. Here, the product is equivalent to the parallel action of individual quantum gates on individual qubits.  Simple application of Formula \eqref{eq:kron} shows that
\begin{align}\label{eq:twoqubits}
\H \otimes \H = \frac{1}{2} \left[\begin{array}{rrrr}
1 & 1 & 1 & 1 \\
1 &  -1& 1 & -1 \\
1 & 1 & -1 & -1 \\
1 & -1 & -1 & 1
\end{array}\right] \quad \mbox{and} \quad \ket{0}\otimes\ket{0}=\begin{bmatrix}
1 \\ 0
\end{bmatrix} \otimes \begin{bmatrix}
1 \\ 0
\end{bmatrix} = \begin{bmatrix}
1 \\ 0 \\ 0 \\ 0
\end{bmatrix} \, .
\end{align} 
We may also denote the left product $\H^{\otimes2}$ and the right product $\ket{00}$ or $\ket{0}\ket{0}$. One may therefore express Equation \eqref{eq:quantparallel} as a series of transformations applied to the 4-vector on the very right. Letting $\ket{10}$, $\ket{01}$ and $\ket{11}$ take on analogous meanings to $\ket{00}$, an immediate result of Equation \eqref{eq:twoqubits} is that
\begin{align}\label{eq:twodmult}
\H^{\otimes2} \ket{00} = \frac{1}{2}\Big(\ket{00} + \ket{01} + \ket{10} + \ket{11}\Big) \, .
\end{align}
The action of $\H^{\otimes2}$ transforms $\ket{00}$ into a superposition of the states $\ket{00}$, $\ket{10}$, $\ket{01}$ and $\ket{11}$, where the probability of selecting each is a uniform $(1/2)^2=1/4$. Writing so many $0$s and $1$s can be tedious, so an alternative notation becomes preferable. Exchanging $\ket{0}$ for $\ket{00}$, $\ket{1}$ for $\ket{01}$, $\ket{2}$ for $\ket{10}$, and $\ket{3}$ for $\ket{11}$, Equation \eqref{eq:twodmult} becomes the slightly more succinct
\begin{align*}
\H^{\otimes2} \ket{0} = \frac{1}{2}\Big(\ket{0} + \ket{1} + \ket{2} + \ket{3}\Big) \, .
\end{align*}
This formula extends generally to operations over $n$ qubits.  Now letting $N=2^n$, we have
\begin{align}\label{eq:superpos}
\H^{\otimes n} \ket{0} = \frac{1}{\sqrt{N}}\Big(\ket{0} + \ket{1} + \dots + \ket{N-1}\Big) =  \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \, ,
\end{align}
and the many-qubit analogue for Equation \eqref{eq:quantparallel} is
\begin{align}\label{eq:quantparallel2}
\ket{0}^{\otimes n}\ket{0} \longrightarrow \left(\frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \right) \ket{0}  \longrightarrow \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \ket{f(x)} \, .
\end{align}
In the next section, we review established quantum algorithms that leverage such many-qubit quantum parallelism for search and minimization.

\section{Quantum search and quantum minimization}

@@
